The Noisy Work of Uncertainty Visualisation

### 2024-01-26 paper.qmd
finished the taxonomy summary bit and added the drawing

---
title: "Uncertainty Literature Review"
author: Harriet Mason
bibliography: references.bib
date: last-modified
format: pdf
editor_options: 
  chunk_output_type: console
---

@@

@@

*The taxonomy* we will use in this paper may seem complicated and difficult to grasp at first glance, but it essentially boils down to three dimensions to consider.Uncertainty is a broad and complicated topic and while oversimplified taxonomies might help us hold the concept of uncertainty in our head, they do little to help with the practical question of how to quantify, visualise, or communicate uncertainty.
- also makes it easier for people to communicate exactly what they need visualisation tools for. Look at geography papers and how their specific uncertainty considerations have been ignored for years.
- Additionally, this definition of uncertainty is helpful in understanding *how* uncertainty should be communicated and the consideration we should make.
- can organise other "uncertainties" according to this framework

*The location* axis of the taxonomy lines up neatly with the typical data analysis pipeline and its importance is backed up by a several philisophical perspectives on statistics as well as discussions on the practical needs of uncertainty visualisation. 
- The usually uncertainty is only considered at the final stage but it needs to be considered through all stages. The big data paradox shrinks confidence intervals but magnifies bias [@Bradley2021] because the confidence intervals shrink due to the sample size but the imprecise target group creates bias.
- @Meng2021 highlights that data needs to be treated as an output from another process rahter than an value neutral input.
- If the distinction between noise and uncertainty is largely task dependent, considerations of quantifying and expressing uncertainty should be considered for the entire duraction of a project; another concept that has been independently identified by several authors [@Kinkeldey2014; @Refsgaard2007]
- While quantification of the uncertainty that arrises from statistical philosophy and data provenance are beyond the scope of this paper, how this uncertainty is communicated is not. The ontological process discussed by @Otsuka2023 where information is tossed as we boil our information down into statistical objects requires us to map our statistical objects back to real world entities. This means the noise introduced and signal lost in earlier stages of the data analysis need to be identified, quantified and expressed in terms of real world outcomes. Visualisation is often the final stage of larger data science project and the down stream effects of poor abstraction and problem at the lower levels become impossible to ignore [@Munzner2009].

*The level* axis forces us to identify how quantifiable the uncertainty is. 
- it also gives us rules for combining uncertainty since statistical uncertainty can be combined with staitstical uncertainty, and outcome uncertainty can be combined with outcome uncertainty. 
- Can discuss each uncertainty in isolation using distributions, ranges, and lists of alternatives, but the total cumulative uncertainty from every stage is more difficult (near impossible) @Spiegelhalter2017
- Approaches to communicating limited understanding include 1) explicit model uncertainty 2) Qualitative scales expressed as strength of evidence (e.g. high to low certainty of evidence) 3) Acknowledge limitations and their possible impact 4) Acknowledged ignorance @Spiegelhalter2017

*The nature axis* allows us to easily communicate if the uncertainty can be reduced or if it needs to be accepted. While this level if not of particular importance to the visualisation of uncertainty, that is not the only purpose of the taxonomy. The nature of the uncertainty is an important element in expressing uncertainty, especually to stakeholders, and connecting the nature to the location makes it easier to understand *how* uncertainty can be reduced.

### 2.2.2 Taxonomies of Visual Uncertainty
There are some taxonomies of visual uncertainty. These taxonmies seem to seek to define "uncertainty visualisation" and a combined term rather than as two separate words. We largely disagree with this conceptualization of uncertainty visualisations as it blurs the line between the mathematical estimations and assumptions (the uncertainty part) and the visual depiction of those mathematical objects (the visualisation part). While we will not focus on these taxonomies, it is still important to mention them and highlight the reasons they are not good practice.

There are a small handful of taxonomies (or typologies) that are specifically for uncertainty visualisations. @Kinkeldey2014 identified give categories of uncertainty based on the work of other literature reivews of uncertainty visualisation. The five categories are: (1) explicit/implicit (directly mapping or showing multiple outcomes) (2) intrinsic/extrinsic (using existing symbols e.g. colour value, or new objects e.g. grids) (3) visually integral/separable (can or cannot be separated from the data) (4) coincidence/adjacent (if data and uncertainty are in integrated or separate views (5) static/dynamic (animation/interaction). Of the groups they identified, they actually only used (4), (2), and (5), left out (1) because most visualisations are explicit, and (3) corresponds to (1) in most cases [@Kinkeldey2014].

@@

While mathematical information *should* be considered the starting point in considering if two visualisations are equivalent, it is not the only consideration. A large amount of work in perceptual tasks, attention, and psychology surrounding charts shows that all the information depicted is not paid attention to in equal measure. Elements such as colour can have sizable impacts on the way a graphic is percieved. Mathematics identifies the information that is *shown* while psychology identifies the difference in information *recieved*. A failure to understand this distinction leads to graphics that are different not only in the way information is depicted but in the information itself. This problem is pervasive in uncertainty visualisation literature.
# 4.?? Noise vs Signal (question +)
The first issue that arises with uncertainty research is that uncertainty is frequently treated as a signal, not noise. This discrepancy is also noted by @Kinkeldey2014 who, in their own literature review on uncertainty methods, comment that the literature seems unsure if uncertainty is simply another variable or if it should be treated differently. This is a fair question to ask. If uncertainty is simply a signal we are trying to convey, why should it be treated any different to any other signal in a visualisation?

>>>>

---
title: "The Noisy Work of Uncertainty Visualisation"
author: Harriet Mason
bibliography: references.bib
date: last-modified
format: pdf
editor_options: 
  chunk_output_type: console
---

@@

*I need to fix the taxonomy illustration because inputs and model uncertainty need to be swapped*

@@

*The taxonomy* we will use in this paper may seem complicated and difficult to grasp at first glance, but it essentially boils down to three dimensions to consider. Uncertainty is a broad and complicated topic and while oversimplified taxonomies might help us hold the concept of uncertainty in our head, they do little to help with the practical question of how to quantify, visualise, or communicate uncertainty. Organising uncertainty according to its location, level and nature makes allows us to consider uncertainty in the miriad of ways it can arise in a data analysis project and better understand *how* this uncertainty should be communicated. With a detailed understanding of the sources of uncertainty and the ways it can be quantified, we can better understand the gaps in the current set of visualisation tools and assess if these gaps need to be filled.

*The location* axis of the taxonomy lines up neatly with the typical data analysis pipeline. Multiple authors in the uncertainty literature have commented on the need to consider quantifying and expressing uncertainty at every stage of a project [@Kinkeldey2014; @Refsgaard2007]. @fig-loc shows the connection between the typical data analysis pipeline and the location element of the uncertainty taxonomy, along with examples of how uncertainty could arise from each of these steps. The relevance of every stage of the uncertainty taxonomy is backed up by both theoretical and practical work in statistics and adjacent fields. @Munzner2009, in her nested model for visualisation, comments on how even something as low level as poor problem abstraction can cause down stream effects and become impossible to ignore. @Otsuka2023 highlights that the choices made in the ontological process of boiling real world events down into statistical objects can influence later stages of analysis due to what was considred signal and what was considered noise. @Meng2021 stresses the importance of treating data as an input from another process with noise and variance, rather than as a neutral input. High level stages such as model selection and estimation have obvious introductions of uncertainty through the choice of models and nature of estimation [^I'm not sure if I need a citation here because this feels obvious]. Location is also acknowledged as an important element of consideration in the taxonomies of several visualsiation papers [^ I will probably need a citation for which ones].

[Illustration of the connection between the data generating process and the location of the uncertainty analysis. Several examples are collected from @utypo and @Munzner2009](location.png){#fig-loc}

*The level* axis forces us to identify how quantifiable the uncertainty is. This element is important because establishing how quantifiable uncertainty is informs us how it can be communicated, something that @Spiegelhalter2017 also identified in the form of "precision". Statistical uncertainty can be quantified, so it can be expressed using a PDF, a variance, an error estimate, or something similar while outcome uncertainty cant be quantified but each scenario may be simulated. Acknowledged ignorance can only be stated in the form of an assumption. Visualisations of climate change scenario uncertainty typically combine the outcome uncertainty of carbon emissions with the and statistical uncertainty of the noisy measurement and modelling system to create a cohesive illustration of the uncertainty surrounding climate change. It is commonly noted that each uncertainty must be discussed in isolation, but combining the uncertainty from every stage is near impossible [@Spiegelhalter2017]. The level axis and climate change visualisations provide tools for how to consider combining these uncertainties. Uncertainties that are the same level but from multiple locations can likely be combined to establish an overall model uncertainty. While uncertainties at different levels cannot be combined, they can still be visualised simultaneously as shown in the climate scenario uncertainty visualisations.

*The nature axis* allows us to easily communicate if the uncertainty can be reduced or if it needs to be accepted. While this level if not of particular importance to the visualisation of uncertainty, that is not the only purpose of the taxonomy. The nature of the uncertainty is an important element in expressing uncertainty, especially to stakeholders, and connecting the nature to the location makes it easier to understand *how* uncertainty can be reduced.

Mapping uncertainty into the 3D space defined by location, level, and nature can expand beyond the narrow case of statistical modelling it was defined for. *@___* included "consensus" as a taxonomy element, however it was not included in the final definition, but it can easily be added as a location that would align with the communication stage of the data pipeline. Several authors [*@_+_@Spiegelhalter2017*] considered the method by which uncertainty was quantified (e.g. by a single value, a function, simulated outcomes, etc), which can be be thought of as a more precise categorisation of the level axis. This flexibility means that the use of this framework is not limited by a specific case and should be used going forward to define what aspect of the uncertainty space a contribution exists in.

*Should I add in a bit where uncertainty papers are organised according to which uncertainties they visualise or should I leave that until the lit review section?*

### 2.2.2 Taxonomies of Visual Uncertainty
There are some taxonomies of visual uncertainty. These taxonomies seem to seek to define "uncertainty visualisation" and a combined term rather than as two separate words. We largely disagree with this conceptualization of uncertainty visualisations as it blurs the line between the mathematical estimations and assumptions (the uncertainty part) and the visual depiction of those mathematical objects (the visualisation part). While we will not focus on these taxonomies, it is still important to mention them and highlight the reasons they are not good practice.

There are a small handful of taxonomies (or typologies) that are specifically for uncertainty visualisations. @Kinkeldey2014 identified give categories of uncertainty based on the work of other literature reviews of uncertainty visualisation. The five categories are: (1) explicit/implicit (directly mapping or showing multiple outcomes) (2) intrinsic/extrinsic (using existing symbols e.g. colour value, or new objects e.g. grids) (3) visually integral/separable (can or cannot be separated from the data) (4) coincidence/adjacent (if data and uncertainty are in integrated or separate views (5) static/dynamic (animation/interaction). Of the groups they identified, they actually only used (4), (2), and (5), left out (1) because most visualisations are explicit, and (3) corresponds to (1) in most cases [@Kinkeldey2014].

@@

While mathematical information *should* be considered the starting point in considering if two visualisations are equivalent, it is not the only consideration. A large amount of work in perceptual tasks, attention, and psychology surrounding charts shows that all the information depicted is not paid attention to in equal measure. Elements such as colour can have sizable impacts on the way a graphic is percieved. Mathematics identifies the information that is *shown* while psychology identifies the difference in information *recieved*. A failure to understand this distinction leads to graphics that are different not only in the way information is depicted but in the information itself. This problem is pervasive in uncertainty visualisation literature.

- The big data paradox shrinks confidence intervals but magnifies bias [@Bradley2021] because the confidence intervals shrink due to the sample size but the imprecise target group creates bias.

# 4.?? Noise vs Signal (question +)
The first issue that arises with uncertainty research is that uncertainty is frequently treated as a signal, not noise. This discrepancy is also noted by @Kinkeldey2014 who, in their own literature review on uncertainty methods, comment that the literature seems unsure if uncertainty is simply another variable or if it should be treated differently. This is a fair question to ask. If uncertainty is simply a signal we are trying to convey, why should it be treated any different to any other signal in a visualisation?
